---
title: "Visualization and Modeling of AirSage Trips Data"
author: "Huajie Yang, Wei Shi, Yi Wang, and Liming Wang (Faculty Advisor)"
date: "12/15/2016"
output: html_document
bibliography: references.bib
---

## Abstract

## Introduction
1. Intro of AirSage Data

2. Describe what we have done

This paper is available on [github](https://cities-lab.github.io/TRB2017DataContest/). We follow the practice of reproducible research [@Gandrud2015] for our work and our writing-up is fully reproducible from a Rmarkdown file and R scripts, the source of which is available on our github repository at https://github.com/cities-lab/TRB2017DataContest (the data files are withheld according to AirSage's confidentiality agreement).


```{r setup, include=FALSE}
if (!require("pacman")) {install.packages("pacman"); library(pacman)}
p_load(dplyr, gridExtra, ggplot2, knitr, printr, scales, magrittr, 
       stargazer, readr, pander, stringr, ggmap, maptools)
p_load_gh("rstudio/leaflet", "dgrtwo/gganimate")

#opts_knit$set(root.dir = here(".."))
knitr::opts_knit$set(root.dir = normalizePath("..")) 
opts_chunk$set(echo=FALSE, message=FALSE, warning=F)
options(scipen=100)
options(digits=3)
```

```{r source, include=F}
source("code/functions.R")
source("code/load_data.R")
source("code/agg_sld.R")
```

## Visualization
### Trip Purpose
```{r}
trips.date_hour_purpose <- trips.df %>% 
  #filter(Start_Weekday) %>% 
  group_by(Start_Date, Start_Hour, Purpose) %>% 
  summarize(trips=sum(Count),
            Start_Day=first(Start_Day),
            Start_Weekend=first(Start_Weekend)) %>% 
  I(#trips_scaled=rescale(trips),
         #Start_Day=format(Start_Date, "%a"),
         #Start_Weekend=ifelse(Start_Day %in% c("Sun", "Sat"), "Weekend", "Weekday"),
         #Start_Day=factor(Start_Day, levels=c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat"))
         ) %>% 
  ungroup()

ggplot(data=trips.date_hour_purpose, aes(x=Start_Hour, y=trips, colour=Purpose, group=Purpose)) +
  #geom_area(position = 'stack') +
  stat_smooth() + ylim(0, 500000) +
  facet_wrap(~Start_Day)
  #scale_y_continuous(labels = percent) +
  #labs(x="standard dev of residuals", y="Normalized mean error") +
  #scale_colour_discrete(guide=FALSE) +
  #scale_y_log10()

ggplot(data=trips.date_hour_purpose, aes(x=Start_Hour, y=trips, colour=Purpose, group=Purpose)) +
  #geom_area(position = 'stack') +
  stat_smooth() + ylim(0, 500000) +
  facet_wrap(~Start_Weekend)

trips.date_hour_purpose %<>% 
  group_by(Start_Date, Purpose) %>% 
  mutate(trips_scaled=scale(trips))

base_size <- 9
(p <- ggplot(trips.date_hour_purpose, aes(Start_Hour, Start_Date)) + geom_tile(aes(fill = trips_scaled),
     colour = "white") + scale_fill_gradient(low = "white", high = "red") + 
     # theme_grey(base_size = base_size) + labs(x = "", y = "") + 
     # scale_x_discrete(expand = c(0, 0)) + scale_y_discrete(expand = c(0, 0)) + 
     # theme(legend.position = "none", axis.ticks = element_blank(), 
     #      axis.text.x = element_text(size = base_size * 0.8, angle = 330, 
     #                               hjust = 0, colour = "grey50"))    +
    facet_wrap(~Purpose)
    )
```

### Subscriber Class
```{r}
trips.date_hour_class <- trips.df %>% 
  #filter(Start_Weekday) %>% 
  group_by(Start_Date, Start_Hour, Subscriber_Class) %>% 
  summarize(trips=sum(Count)) %>% 
  mutate(trips_scaled=rescale(trips),
         Start_Day=format(Start_Date, "%a"),
         Start_Weekend=ifelse(Start_Day %in% c("Sun", "Sat"), "Weekend", "Weekday"),
         Start_Day=factor(Start_Day, levels=c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat"))
         ) %>% 
  ungroup()

ggplot(data=trips.date_hour_class, aes(x=Start_Hour, y=trips, colour=Subscriber_Class, group=Subscriber_Class)) +
  #geom_area(position = 'stack') +
  stat_smooth() + ylim(0, 500000) +
  facet_wrap(~Start_Day)
  #scale_y_continuous(labels = percent) +
  #labs(x="standard dev of residuals", y="Normalized mean error") +
  #scale_colour_discrete(guide=FALSE) +
  #scale_y_log10()

ggplot(data=trips.date_hour_class, aes(x=Start_Hour, y=trips, colour=Subscriber_Class, group=Subscriber_Class)) +
  #geom_area(position = 'stack') +
  stat_smooth() + ylim(0, 500000) +
  facet_wrap(~Start_Weekend)

trips.date_hour_class %<>% 
  group_by(Start_Date, Subscriber_Class) %>% 
  mutate(trips_scaled=scale(trips))
```

Comparing the travel pattern differences among subscriber classes, we found that short term visitors made the most trips, followed by long term visitors, resident workers and home workers, whereas there were relatively less trips made by inbound and outbound commuters. In terms of time of day, the trip count was an apparent inverse U curve indicating more trips were made in daytime, with PM peak trend for resident workers. All subscriber classes seemed to have late starts in the mornings and early ends in the evenings in weekends than weekdays. Resident workers made relatively more trips in late afternoon time, possibly after-work errand trips, in the weekdays than weekends.

As there were significant differences between weekdays and weekends, we closely compare the travel pattern differences between weekdays and weekends among different subscriber classes. It was clear that almost all subscriber class, except long term visitors made more trips in weekdays than in weekends. The resident workers' travel patterns of more late afternoon trips in weekdays was more apparent in the weekend/weekday comparison graph. In addition, the variation of trip generation, represented by the grey shady areas around the lines, were bigger in weekends than weekdays, indicating trips generation in weekends were less predictable than weekdays.


```{r}
(p <- ggplot(trips.date_hour_class, aes(Start_Hour, Start_Date)) + geom_tile(aes(fill = trips_scaled),
     colour = "white") + scale_fill_gradient(low = "white", high = "red") + 
    facet_wrap(~Subscriber_Class)
    )
```

```{r animation}
# first compute culumative trips
trips.d1 <- trips.df %>% 
  filter(Start_Date=="2014-04-01")

cumtrips.out <- trips.d1 %>%   
  group_by(Origin_Zone, Start_Hour) %>% 
  summarize(trips=sum(Count)) %>% 
  mutate(cumtrips.out=cumsum(trips)) %>% 
  select(Origin_Zone, Start_Hour, cumtrips.out)

cumtrips.in <- trips.d1 %>%   
  group_by(Destination_Zone, Start_Hour) %>% 
  summarize(trips=sum(Count)) %>% 
  mutate(cumtrips.in=cumsum(trips)) %>% 
  select(Destination_Zone, Start_Hour, cumtrips.in)
  
cumtrips <- cumtrips.in %>% 
  left_join(cumtrips.out, 
            by=c("Destination_Zone"="Origin_Zone", "Start_Hour"="Start_Hour")) %>% 
  mutate(cumtrips=cumtrips.in - cumtrips.out,
         cumtrips=ifelse(cumtrips<0, 0, cumtrips),
         id=as.character(Destination_Zone))

spdf.cumtrips <- spdf.ff %>% 
  left_join(cumtrips, by=c("id"="id"))

map <- ggplot() + geom_polygon(aes(x = long, y = lat, group=group, fill=cumtrips),
                               data = spdf.cumtrips, color ="black", 
                               size = 0) +
  coord_map()+
  scale_fill_distiller(name="trips arrived", palette = "RdBu", breaks = pretty_breaks(n = 5))+
  theme_nothing(legend = TRUE)
map

require(gganimate)
theme_set(theme_bw())
p <- ggplot() + geom_polygon(aes(x = long, y = lat, group=group, fill=cumtrips, frame=Start_Hour),
                               data = spdf.cumtrips, color ="black", 
                               size = 0) +
    coord_map()+
    scale_fill_distiller(name="trips arrived", palette = "RdBu", breaks = pretty_breaks(n = 5))
gg_animate(p)
```

```{r ggmaps}


shpfile <- 'data/taz/OUATS_AirSage_TAZ.shp'
spdf <- read.shpfile(shpfile)

spdf.ff <- fortify(spdf, region="TAZ_ID")
spdf.ff %<>% left_join(m0401.o, by=c("id"="TAZ_ID"))

map <- ggplot() + geom_polygon(aes(x = long, y = lat, group=group, fill=trips),
                               data = spdf.ff, color ="black", 
                               size = 0) +
  coord_map()+
  scale_fill_distiller(name="trips originated", palette = "RdBu", breaks = pretty_breaks(n = 5))+
  theme_nothing(legend = TRUE)
map
```


```{r leaflet_maps}
spdf@data %<>% left_join(m0401.o %>% mutate(TAZ_ID=as.integer(TAZ_ID)))
pal.qtl <- colorQuantile(palette = "YlOrRd", domain = spdf@data$trips)
pal.num <- colorNumeric (palette = "YlOrRd", domain = spdf@data$trips)
leaflet(spdf) %>%
  addPolygons(
    stroke = FALSE, fillOpacity = 0.5, smoothFactor = 0.5,
    fillColor = ~pal.qtl(trips),
    label=~stringr::str_c(
                TAZ_ID,' ',
                formatC(trips, big.mark = ',', format='d')),
    labelOptions= labelOptions(direction = 'auto'),
    highlightOptions = highlightOptions(
                color='#ff0000', opacity = 1, weight = 2, fillOpacity = 1,
                bringToFront = TRUE, sendToBack = TRUE)
  )
```

## References
<! -- Gandrud, C., 2015. Reproducible Research with R and R Studio, Second Edition, 2 edition. ed. Chapman and Hall/CRC, Boca Raton. -->


<!--## Session Info-->
```{r session_info}
#pander(devtools::session_info())
```